---
title: "The diffusion Model"
---

```{r}
rm(list = ls())
library(tidyverse)
library(dfoptim)
library(rtdists)
library(rstan)
library(bayesplot)
```

We can write down Equation 5 from [Bogacz 2006 paper](https://pubmed.ncbi.nlm.nih.gov/17014301/) to simulate the process described by the Diffusion Model *without* across-trial variability:

```{r}
dm_path <- function(drift, threshold, ndt, rel_sp=.5, noise_constant=1, dt=0.001, max_rt=10) {
  max_tsteps <- max_rt/dt
  
  # initialize the diffusion process
  tstep <- 0
  x <- c(rel_sp*threshold) # vector of accumulated evidence at t=tstep
  time <- c(ndt)
  
  # start accumulating
  while (0 < x[tstep+1] & x[tstep+1] < threshold & tstep < max_tsteps) {
    x <- c(x, x[tstep+1] + rnorm(mean=drift*dt, sd=noise_constant*sqrt(dt), n=1))
    time <- c(time, dt*tstep + ndt)
    tstep <- tstep + 1
  }
  return (data.frame(time=time, dv=x))
}
```

And visualize it:
```{r echo=TRUE, message=FALSE, warning=FALSE}
gen_drift = .3
gen_threshold = 1
gen_ndt = .23
sim_path <- dm_path(gen_drift, gen_threshold, gen_ndt)

ggplot(data = sim_path, aes(x = time, y = dv))+
  geom_line(size = .5) +
  geom_hline(yintercept=gen_threshold, size=1.5) +
  geom_hline(yintercept=0, size=1.5)
```

To have a look at the whole distribution, though, we want to simulate more trials:

```{r}
random_dm <- function(n_trials, drift, threshold, ndt, rel_sp=.5, noise_constant=1, dt=0.001, max_rt=10) {

  acc <- rep(NA, n_trials)
  rt <- rep(NA, n_trials)
  max_tsteps <- max_rt/dt

  # initialize the diffusion process
  tstep <- 0
  x <- rep(rel_sp*threshold, n_trials) # vector of accumulated evidence at t=tstep
  ongoing <- rep(TRUE, n_trials) # have the accumulators reached the bound?

  # start accumulating
  while (sum(ongoing) > 0 & tstep < max_tsteps) {
    x[ongoing] <- x[ongoing] + rnorm(mean=drift*dt,
                                     sd=noise_constant*sqrt(dt),
                                     n=sum(ongoing))
    tstep <- tstep + 1
    
    # ended trials
    ended_correct <- (x >= threshold)
    ended_incorrect <- (x <= 0)
    
    # store results and filter out ended trials
    if(sum(ended_correct) > 0) {
      acc[ended_correct & ongoing] <- 1
      rt[ended_correct & ongoing] <- dt*tstep + ndt
      ongoing[ended_correct] <- FALSE
    }
    
    if(sum(ended_incorrect) > 0) {
      acc[ended_incorrect & ongoing] <- 0
      rt[ended_incorrect & ongoing] <- dt*tstep + ndt
      ongoing[ended_incorrect] <- FALSE
    }
  }
  return (data.frame(trial=seq(1, n_trials), accuracy=acc, rt=rt))
}
```

And have a look at the average performance and shape of the RT distributions:

```{r echo=TRUE, message=FALSE, warning=FALSE}
sim_data <- random_dm(n_trials=1000, drift=.7, threshold=1.5, ndt=.23)
summary(sim_data)

ggplot(data = sim_data, mapping = aes(x = rt, fill = factor(accuracy))) + 
  geom_histogram(binwidth=.05, alpha = .3, position="identity")
```

The same result can be achieved with the `rdiffusion` function of the `rtdists`package:

```{r echo=TRUE, message=FALSE, warning=FALSE}
sim_data2 <- rdiffusion(n=1000, a=1.5, v=.7, t0=.23)
sim_data2$accuracy = 0
sim_data2[sim_data2$response=="upper", "accuracy"] = 1
summary(sim_data2)
```

## Parameter recovery with MLE

An efficient approximation to the likelihood function of the DM can be found in the [Navarro & Fuss paper](https://www.sciencedirect.com/science/article/pii/S0022249609000200?via%3Dihub#appC). In the appendix, you can find the matlab code and convert it to R. However, a computationally efficient version of it can be also found in the `rtdists` package as well. We just need to wrap it to use in the MLE fitting later on:

```{r echo=TRUE, message=FALSE, warning=FALSE}
log_likelihood_dm <- function(par, data, ll_threshold=1e-10) {
  
  density <- ddiffusion(rt=data$rt, response=data$response, a=par[1], v=par[2], t0=par[3])
  
  density[density <= ll_threshold] = ll_threshold # put a threhsold on very low likelihoods for computability
  
  return(sum(log(density)))
}
```

We can thus use the Nelder-Mead algorithm in the `dfoptim` package to estimate the parameters:

```{r echo=TRUE, message=FALSE, warning=FALSE}
starting_values = c(.5, 1, .1) # set some starting values

print(log_likelihood_dm(starting_values, data=sim_data2)) # check that starting values are plausible

fit1 <- nmkb(par = starting_values,
             fn = function (x) log_likelihood_dm(x, data=sim_data2),
             lower = c(0, -10, 0),
             upper = c(10, 10, 5),
             control = list(maximize = TRUE))
print(fit1$par) # print estimated parameters
```

## Parameter Recovery with stan

We can also recover the generating parameters of the simulated data with stan, to assess th model's identifialbility.

First, we need to prepare our data for stan:

```{r message=FALSE, warning=FALSE}
sim_data$accuracy_recoded = sim_data$accuracy
sim_data[sim_data$accuracy==0, "accuracy_recoded"] = -1

sim_data_for_stan = list(
  N = dim(sim_data)[1],
  accuracy = sim_data$accuracy_recoded,
  rt = sim_data$rt,
  starting_point = 0.5
)
```

And then we can fit the model:

```{r message=FALSE, warning=FALSE}
fit1 <- stan(
  file = "stan_models/DM.stan",                      # Stan program
  data = sim_data_for_stan,                          # named list of data
  chains = 2,                                        # number of Markov chains
  warmup = 1000,                                     # number of warmup iterations per chain
  iter = 3000,                                       # total number of iterations per chain
  cores = 2                                          # number of cores (could use one per chain)
)
```

Compare the generating parameters with the recovered ones and check for convergence looking at the Rhat measures:

```{r echo=TRUE, message=FALSE, warning=FALSE}
print(fit1, pars = c("drift", "transf_threshold", "transf_ndt"))
```

And (visually) assess the model's convergence as well as some more sampling diagnostics:

```{r echo=TRUE, message=FALSE, warning=FALSE}
traceplot(fit1, pars = c("drift", "transf_threshold", "transf_ndt"), inc_warmup = FALSE, nrow = 2)

sampler_params <- get_sampler_params(fit1, inc_warmup = TRUE)
summary(do.call(rbind, sampler_params), digits = 2)
```

More plotting:
```{r}
posterior <- as.matrix(fit1)

plot_title <- ggtitle("Posterior distributions",
                      "with medians and 95% intervals")
mcmc_areas(posterior,
           pars = c("drift", "transf_threshold", "transf_ndt"),
           prob = 0.95) + plot_title
```
