---
title: "Fontanesi and colleagues 2019"
---

```{r echo=TRUE, message=FALSE, warning=FALSE}
rm(list = ls())
library(tidyverse)
library(dfoptim)
library(rtdists)
library(rstan)
library(bayesplot)
```

First, we can load the data of [our study from 2019](https://link.springer.com/article/10.3758/s13423-018-1554-2):

```{r echo=TRUE, message=FALSE, warning=FALSE}
data <- read.csv('data/fontanesi2019.csv')
data <- select(data, -X) # drop pandas index column
data <- data[data$participant < 6,] # select only 5 participants otherwise it takes too long
# add accuracy for stan data
data$accuracy_recoded <- data$accuracy
data[data$accuracy==0, "accuracy_recoded"] <- -1

# add conditions based on choice pairs
data[(data$inc_option == 1)&(data$cor_option == 2),"condition_label"] <- "difficult_low"
data[(data$inc_option == 1)&(data$cor_option == 3),"condition_label"] <- "easy_low"
data[(data$inc_option == 2)&(data$cor_option == 4),"condition_label"] <- "easy_high"
data[(data$inc_option == 3)&(data$cor_option == 4),"condition_label"] <- "difficult_high"

data[(data$inc_option == 1)&(data$cor_option == 2),"condition"] <- 1
data[(data$inc_option == 1)&(data$cor_option == 3),"condition"] <- 2
data[(data$inc_option == 2)&(data$cor_option == 4),"condition"] <- 3
data[(data$inc_option == 3)&(data$cor_option == 4),"condition"] <- 4
```

The first model we can try to fit, is a diffusion model (DM) in which we vary drift-rate and threshold based on the condition (or choice pair). The DM assumptions are actually violated in this task: when we fit the DM without learning component, we assume that the trials are somewhat independent. While this is rarely the case even in non-learning tasks (e.g., because of post-error effects, practice effects) this is clearly not the case in learning tasks. However, this can give us a good idea of context-dependent effects on the RT distributions such as the ones we observed in the following plots:

```{r echo=TRUE, message=FALSE, warning=FALSE}
ggplot(data = data, aes(x = condition, y = accuracy, color=condition_label))+
  stat_summary(fun = "mean", geom="point",  size=4) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar",  size=.5, width=.3)

ggplot(data = data, aes(x = condition, y = rt, color=condition_label))+
  stat_summary(fun = "mean", geom="point",  size=4) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar",  size=.5, width=.3)
```

Now we can prepare the data for `stan`. Note that this is a bit different for hierarchical models, and also when we want to fit separate parameters per condition:

```{r echo=TRUE, message=FALSE, warning=FALSE}
unique(data$participant)

sim_data_for_stan <- list(
  N = dim(data)[1],
  L = length(unique(data$participant)),
  C = 4,
  condition = data$condition,
  participant = data$participant,
  accuracy = data$accuracy_recoded,
  rt = data$rt,
  starting_point = 0.5
)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
fit1 <- stan(
  file = "stan_models/hierDDM_cond.stan",            # Stan program
  data = sim_data_for_stan,                          # named list of data
  chains = 2,                                        # number of Markov chains
  warmup = 1000,                                     # number of warmup iterations per chain
  iter = 3000,                                       # total number of iterations per chain
  cores = 2                                          # number of cores (could use one per chain)
)
```

We can summarize the group parameters:
```{r echo=TRUE, message=FALSE, warning=FALSE}
print(fit1, pars = c("mu_drift[1]", "mu_drift[2]", "mu_drift[3]", "mu_drift[4]",
                     "sd_drift[1]", "sd_drift[2]", "sd_drift[3]", "sd_drift[4]",
                     "mu_threshold[1]", "mu_threshold[2]", "mu_threshold[3]", "mu_threshold[4]",
                     "sd_threshold[1]", "sd_threshold[2]", "sd_threshold[3]", "sd_threshold[4]",
                     "mu_ndt", "sd_ndt"))
```

Compare the group mean of the drift-rates across conditions:
```{r echo=TRUE, message=FALSE, warning=FALSE}
posterior <- as.matrix(fit1)

plot_title <- ggtitle("Posterior distributions mu drift-rate per condition",
                      "with medians and 95% intervals")
mcmc_areas(posterior,
           pars = c("mu_drift[1]", "mu_drift[2]", "mu_drift[3]", "mu_drift[4]"),
           prob = 0.95) + plot_title
```

As well as the thresholds:
```{r echo=TRUE, message=FALSE, warning=FALSE}
plot_title <- ggtitle("Posterior distributions mu threshold per condition",
                      "with medians and 95% intervals")

mcmc_areas(posterior,
           pars = c("mu_threshold[1]", "mu_threshold[2]", "mu_threshold[3]", "mu_threshold[4]"),
           prob = 0.95) + plot_title
```

And for example we can compare participants' NDT parameters:
```{r echo=TRUE, message=FALSE, warning=FALSE}
plot_title <- ggtitle("Posterior distributions individual NDT parameters",
                      "with medians and 95% intervals")

mcmc_areas(posterior,
           pars = c("ndt_sbj[1]", "ndt_sbj[2]", "ndt_sbj[3]", "ndt_sbj[4]", "ndt_sbj[5]"),
           prob = 0.95) + plot_title
```

As we expected, the drift-rates for the easier conditions tend to be higher, and the thresholds for the high-value pairs tends to be lower.